---
title: "ISPPD Workshop #2 Evaluating Vaccine Impact using Time Series Data"
author: "Daniel M. Weinberger (daniel.weinberger@yale.edu) and Kayoko Shioda (kayoko.shioda@yale.edu)"
date: "Sunday, April 15, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

First, download "Brazil_acp.csv" and save it in your folder.

Then, run the following section to import the dataset.

```{r}
# Set working directory
setwd("C:/Users/dmw63/Dropbox (Personal)/ISPPD workshop") # Please update this line 

# Import the data in a .csv file
d <- read.csv("Brazil_acp.csv")
```

Let's explore the dataset a little bit...

```{r}
# Explore the dataset
names(d)
head(d)
table(d$age_group)
```

Let's take a look at a date variable. How does it look like? Is it in a right format?

```{r}
class(d$date) # "factor" --> Need to change it to "date"
head(d$date) 

# Change the type of the date variable so that R can recognize it as a date variable
d$date <- as.Date(d$date,format="%m/%d/%Y")
class(d$date) # Now it's changed to "Date"
head(d$date) 
```

Next, let's load packages that we will be using in the following sections.

```{r}
# Load libraries
library(MASS)
library(lubridate)
# If you do not have these packages installed, please run the following line.
# Replace "PackageName" with the name of the package you'd like to install.
#install.packages("PackageName") 
```



## Part 1. Visualize the Data

### Part 1-a. J12-18

Make a plot for the time series for all-cause pneumonia hospitalizations (ICD10 code: J12-18) among children <12 months of age.

Sort the dataset by date, and make the same plots for <12 mo and 80+ yo.
```{r}
# Sort the dataset by date
d <- d[order(d$date),]

# <12 mo
plot(J12_18 ~ date, data=d[d$age_group=="<1",], 
     type="l", bty="l", col="blue", lwd=2, 
     ylim=c(0,max(d$J12_18[d$age_group=="<1"])),
     xlab="Months", ylab="Number of hospitalizations", 
     main="Monthly number of J12-18, <12 mo")
      abline(v=as.Date("2010-01-01"), lty=2)

# 80+ yo
plot(J12_18 ~ date, data=d[d$age_group=="80+",], 
     type="l", bty="l", col="darkgreen", lwd=2,
     ylim=c(0,max(d$J12_18[d$age_group=="80+"])),
     xlab="Months", ylab="Number of hospitalizations", 
     main="Monthly number of J12-18, 80+ yo")
    abline(v=as.Date("2010-01-01"), lty=2)
```

What kind of trend do you see in J12-18 for each age group?


### Part 1-b. ACJ_NOJ

Plot the time series for non-respiratory hospitalizations (i.e., ACH_NOJ) for <12 mo and 80+ yo. This variable will be used as an offset for regression models. 

First, to make the following analyses easier, let's subset the datasets into two age groups (<12 mo and 80+ yo).

```{r}
young <- d[d$age_group=="<1",]
old <- d[d$age_group=="80+",]
```

Now let's make plots for ACH_NOJ.

```{r}
# <12 mo
plot(ACH_NOJ ~ date, data=young, bty="l", type="l",
     ylim=c(0,max(old$ACH_NOJ)),
     xlab="Months", ylab="Number of hospitalizations",
     col="blue", main="ACH_NOJ, <1 yo")
 abline(v=as.Date("2010-01-01"), lty=2)

# 80+ yo
# Version 1 (y axis from zero to the max. number of hospitalizations)
plot(ACH_NOJ ~ date, data=old, bty="l", type="l",
     ylim=c(0,max(old$ACH_NOJ)),
     xlab="Months", ylab="Number of hospitalizations", 
     col="darkgreen", main="ACH_NOJ, 80+ yo")
 abline(v=as.Date("2010-01-01"), lty=2)
# Verson 2 (zoom in)
plot(ACH_NOJ ~ date, data=old, bty="l", type="l",
     xlab="Months", ylab="Number of hospitalizations", 
     col="darkgreen", main="ACH_NOJ, 80+ yo (zoom)")
 abline(v=as.Date("2010-01-01"), lty=2)
```

What kind of trend do you see in ACH_NOJ for each age group?



## Part 2. Negative Binomial Regression

First, we will fit a regession just to the **pre-vaccne** data and will extrapolate the trend to the post-vaccine period to estimate the **counterfactual**. 

Because the outcome is a **COUNT** variable, it is most appropriate to use a log-linked Poisson or negative binomial regression, rather than linear regression.

Due to the over-dispersion present in the data, we will fit a negative binonimal regression.


### Part 2-a. Set up

In order to fit a model just to the pre-vaccine period, set the outcome (J12-18) to missing (NA) for the post vaccine period.

NOTE: PCV10 was introduced on Jan 1, 2010 in Brazil.

```{r}
# Create a new variable J12_18_pre which is NA (missing) in the post-vaccine period.
# <12 mo
young$J12_18_pre <- young$J12_18
young$J12_18_pre[which(young$date>="2010-01-01")] <- NA 
# 80+
old$J12_18_pre <- old$J12_18
old$J12_18_pre[which(old$date>="2010-01-01")] <- NA 

# Check if it was created as we want.
#data.frame(young$date, young$J12_18, young$J12_18_pre)
#data.frame(old$date, old$J12_18, old$J12_18_pre)
```

Next, let's create on offset term for negative binomial regression using ACH_NOJ (in a log scale).

```{r}
# Create an offset term in a log scale--this is the denominator for the regression.
young$log_offset <- log(young$ACH_NOJ)
old$log_offset <- log(old$ACH_NOJ)
```

We will also create a time index variable to control for a long term linear trend.

```{r}
# Create a time index variable (1, 2, 3, 4, ..., number of datapoints)
young$time <- 1:nrow(young)
old$time <- 1:nrow(old)
young$month<-as.factor(month(young$date))
old$month<-as.factor(month(old$date))

```

As the outcome J12-18 shows a clear seasonality, we will adjust for it in the regression model. We can do it in two ways:
  * Using monthly dummy variables  (We will do this here)
  * Using harmonic terms (sine, cosine)
### Part 2-b. Fit a negative binomial model

Fit negative binomial models to the prevaccine data.

```{r}
NB_yng_s1 <- glm.nb(J12_18_pre ~ time + month         + offset(log_offset), data=young)
NB_old_s1 <- glm.nb(J12_18_pre ~ time + month         + offset(log_offset), data=old)
```

Exrapolate the trend to the post-vaccine period and generate the counterfactual for J12-18.

```{r}
Pred_NB_yng_s1 <- predict(NB_yng_s1, newdata=young, type="response", se.fit=T)
#Pred_NB_yng_s2 <- predict(NB_yng_s2, newdata=young, type="response", se.fit=T)
Pred_NB_old_s1 <- predict(NB_old_s1, newdata=old,   type="response", se.fit=T)
#Pred_NB_old_s2 <- predict(NB_old_s2, newdata=old,   type="response", se.fit=T)
```

Plot time series for observed J12-18 vs. counterfactual J12-18.
```{r}
# <12 mo
plot(Pred_NB_yng_s1$fit ~ young$date,
     type="l",col="red", bty="l", lwd=2,
     ylim=c(0,max(c(young$J12_18, Pred_NB_yng_s1$fit))),
     ylab="Number of hospitalizations", xlab="Months", 
     main="Negative binomial model (<12 mo)")
lines(J12_18 ~ date, data=young, col="black",lwd=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2, lwd=2)
legend(x="bottomleft",legend=c("Observed","Monthly dummy","Harmonics"),
       col=c("black","red","blue"),lty=c(1,1,1),bty="n")

# 80+ mo
plot(Pred_NB_old_s1$fit ~ old$date,
     type="l",col="red", bty="l", lwd=2,
     ylim=c(0,max(c(old$J12_18, Pred_NB_old_s1$fit))),
     ylab="Number of hospitalizations", xlab="Months", 
     main="Negative binomial model (80+ yo)")
lines(J12_18 ~ date, data=old, col="black",lwd=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2, lwd=2)
legend(x="bottomleft",legend=c("Observed","Monthly dummy","Harmonics"),
       col=c("black","red","blue"),lty=c(1,1,1),bty="n")
```


### Part 2-c. Rate ratios (RRs)

Calculate the rate ratios (RRs)

```{r}
RR_NB_yng <- young$J12_18/Pred_NB_yng_s1$fit
RR_NB_old <- old$J12_18/Pred_NB_old_s1$fit
```

Plot RRs by time.
```{r}
# <12 mo
plot(RR_NB_yng ~ young$date, type="l", bty="l", col="red",
     main="RR for <12 mo", xlab="Months", ylab="Estimated RR", ylim=c(0.5,1.5))
abline(h=1,col="darkgrey",lty=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2)
# 80+ yo
plot(RR_NB_old ~ old$date, type="l", bty="l", col="red",
     main="RR for 80+ years", xlab="Months", ylab="Estimated RR", ylim=c(0.5,1.5))
abline(h=1,col="darkgrey",lty=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2)
```


### Part 2-e. Leave-one-season-out analysis

As a sensitivity analysis, we can fit a series of negative binomial models by excluding one season at a time. 

For example, the 1st model will be fit to the pre-vaccine data excluding the first year of the pre-vaccine period; the 2nd model will be fit to the pre-vaccine data excluding the second year... 

There are 6 years of pre-vaccine data, so we will fit 6 models.

Let's start with the young age group.

```{r}
# First, let's create an empty matrix to store results.
lvso_yng <- matrix(NA, nrow=nrow(young), ncol=6)
for (i in 1:6) {
  
  # 1. Create J12_18_pre as before
  young$J12_18_pre <- young$J12_18
  young$J12_18_pre[which(young$date>="2010-01-01")] <- NA 
  
  # 2. Exclude one season from the pre-vaccine period
  k <- (12*(i-1)+1):(12*(i-1)+12)
  young$J12_18_pre[k] <- NA 
  
  # 3. Fit a negative binomial model
  NB_yng_lvso <- glm.nb(J12_18_pre ~ time+month+offset(log_offset), data=young)

  # 4. Extrapolate a trend to the post-vaccine period
  Pred_NB_yng_lvso <- predict(NB_yng_lvso, newdata=young, type="response", se.fit=T)
  
  # 5. Save a result in a matrix
  lvso_yng[,i] <- Pred_NB_yng_lvso$fit
}
```

Plot observed vs. counterfactual.

```{r}
plot(J12_18 ~ date, data = young,
     type="l",col="black", bty="l", lwd=2,
     ylim=c(0,max(c(young$J12_18))),
     ylab="Number of hospitalizations", xlab="Months", 
     main="Negative binomial model (<12 mo)")
lines(lvso_yng[,i] ~ young$date, data=young, col="red")
col <- c("blue","orange","darkgreen","purple","darkgrey")
for (i in 2:6) {
  lines(lvso_yng[,i] ~ date, data=young, col=col[i],lwd=2)
}
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2, lwd=2)
```

Calculate and plot RRs.

```{r}
RR_lvso_yng <- young$J12_18/lvso_yng
plot(RR_lvso_yng[,1] ~ young$date, type="l", bty="l", col="red",
     main="RR for <12 mo", xlab="Months", ylab="Estimated RR", ylim=c(0.5,1.5))
for (i in 2:6) {
  lines(RR_lvso_yng[,i] ~ date, data=young, col=col[i],lwd=2)
}
abline(h=1,col="darkgrey",lty=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2)
```




## Part 3. Interrupted Time Series Analysis

In this section, we compare a simple univariate linear regression with interupted time series regressions where we test whether the slope of the line changes after vaccine introduction.

### Part 3-a. Set up

First, let's create the following dummy variables for the post-vaccine period.
   * period1: 1 if 1-12 months after PCV10 introduction
   * period2: 1 if >12 months after PCV10 introduction

```{r}
# <12 mo
young$period1 <- 0
young$period2 <- 0
young$period1[young$date>="2010-01-01" & young$date<"2011-01-01"] <- 1
young$period2[young$date>="2011-01-01"] <- 1
# 80+ yo
old$period1 <- 0
old$period2 <- 0
old$period1[old$date>="2010-01-01" & old$date<"2011-01-01"] <- 1
old$period2[old$date>="2011-01-01"] <- 1
```


### Part 3-b. Fit 3 models

Fit an interrupted time series model as follows.

```{r}
ITS_yng <- glm.nb(J12_18 ~ month + time*period1 + time*period2, data=young)
ITS_old <- glm.nb(J12_18 ~ month + time*period1 + time*period2, data=old)
```

NOTE: This model includes time, period1, and period2 althouth these terms are not explicitly written in the code above. 


### Part 3-c. Plot fitted values

Calculate fitted values.

```{r}
Pred_ITS_yng <- predict(ITS_yng, newdata=young, type="response", se.fit=T)
Pred_ITS_old <- predict(ITS_old, newdata=old,   type="response", se.fit=T)
```

Make plots for the observed vs. fitted.

```{r}
# <12 mo
plot(Pred_ITS_yng$fit ~ young$date,
     type="l",col="red", bty="l", lwd=2,
     ylim=c(0,max(young$J12_18)),
     ylab="Number of hospitalizations", xlab="Months", 
     main="Interrupted Time Series Model (<12 mo)")
lines(J12_18 ~ date, data=young, col="black",lwd=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2, lwd=2)
abline(v=as.Date("2011-01-01"),col="darkgrey",lty=2, lwd=2)
legend("bottomleft", legend=c("Observed","ITS"),col=c("black","red"),lty=c(1,1),bty="n")

# 80+ yo
plot(Pred_ITS_old$fit ~ young$date,
     type="l",col="red", bty="l", lwd=2,
     ylim=c(0,max(old$J12_18)),
     ylab="Number of hospitalizations", xlab="Months", 
     main="Interrupted Time Series Model (80 yo)")
lines(J12_18 ~ date, data=old, col="black",lwd=2)
abline(v=as.Date("2010-01-01"),col="darkgrey",lty=2, lwd=2)
abline(v=as.Date("2011-01-01"),col="darkgrey",lty=2, lwd=2)
legend("bottomright", legend=c("Observed","ITS"),col=c("black","red"),lty=c(1,1),bty="n")
```


### Part 3-d. Generate counterfactual and estimate the impact of PCV10

Calculate the counterfactual which is the number of cases expected without PCV10. In this case, that's the following part of the model:

$$ 
\beta_0 + \beta_1*time + \beta_2*sin12+\beta_3*cos12
$$

```{r}
# < 12mo
#cf_yng <-exp(ITS_yng$coef[1] + ITS_yng$coef[2]*young$sin12 + ITS_yng$coef[3]*young$cos12 + ITS_yng$coef[4]*young$time)
cf_yng<- Pred_ITS_yng$fit / exp(young$period1*ITS_yng$coef['period1'] + young$period2*ITS_yng$coef['period2'] 
                                +young$period1*young$time*ITS_yng$coef['time:period1'] 
                                +young$period2*young$time*ITS_yng$coef['time:period2'] )
# 80+ yo
cf_old<- Pred_ITS_old$fit / exp(old$period1*ITS_old$coef['period1'] +old$period2*ITS_old$coef['period2'] 
                                +old$period1*old$time*ITS_old$coef['time:period1'] 
                                +old$period2*old$time*ITS_old$coef['time:period2'] )
```

Calculate and plot the number of cases averted.

```{r}
# First, let's reformat ITS_###3$fit as follows:
str(Pred_ITS_yng$fit) # It is a "named number", so let's unmane them
Pred_ITS_yng <- unname(Pred_ITS_yng$fit)
Pred_ITS_old <- unname(Pred_ITS_old$fit)

# Calculate the number of cases we averted using our intervention
casesaverted_yng <- Pred_ITS_yng - cf_yng
casesaverted_old <- Pred_ITS_old - cf_old

# Plot
plot(casesaverted_yng, col="orange", main="Cases averted, <12 mo", pch=16)
abline(v=73, col="darkgrey", lty=2)
abline(v=85, col="darkgrey", lty=2)
plot(casesaverted_old, col="orange", main="Cases averted, 80+ yo", pch=16)
abline(v=73, col="darkgrey", lty=2)
abline(v=85, col="darkgrey", lty=2)
```

What about the change in rate?

```{r}
# <12 mo
RR_yng <- Pred_ITS_yng/cf_yng
plot(RR_yng, type="l", col="blue", lwd=2, main="Rate ratio, <12 mo",
     xlab="Months",ylab="Estimated RR", ylim=c(0.5, 1.5))
abline(v=73, col="darkgrey", lty=2)
abline(v=85, col="darkgrey", lty=2)
abline(h=1, col='gray', lty=2)
# 80+
RR_old <- Pred_ITS_old/cf_old
plot(RR_old, type="l", col="blue", lwd=2, main="Rate ratio, 80+ yo",
     xlab="Months",ylab="Estimated RR", ylim=c(0.5, 1.5))
abline(v=73, col="darkgrey", lty=2)
abline(v=85, col="darkgrey", lty=2)
abline(h=1, col='gray', lty=2)

```


Thank you for your participation! Please feel free to contact us anytime if you have any questions! Daniel M. Weinberger (daniel.weinberger@yale.edu) and Kayoko Shioda (kayoko.shioda@yale.edu)